---
title: "riemstats"
author: "Nicolas Escobar, Jaroslaw Harezlak"
date: "2025-02-12"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{riemstats}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

The `riemstats` package provides specialized statistical methods for analyzing brain connectivity matrices (connectomes) from multi-site neuroimaging studies. When working with functional or structural connectivity data represented as symmetric positive definite (SPD) matrices, this package offers both harmonization techniques to address site-specific effects and ANOVA methods that respect the Riemannian geometry of SPD matrices.

## The Multi-Site Connectome Challenge

Modern neuroimaging studies often collect brain connectivity data across multiple sites to increase sample size and generalizability. However, this introduces significant challenges associated to the fact that different scanners, protocols, and populations introduce systematic biases.

Traditional statistical methods fail to address these challenges simultaneously, either ignoring the geometric structure of SPD matrices or lacking appropriate harmonization capabilities.

## Core Capabilities

`riemstats` addresses these challenges through two complementary sets of tools:

1. **Harmonization Methods**: Remove unwanted site/scanner effects while preserving biological signals
  - ComBat harmonization adapted for SPD matrices (see Honorro et al.)
  - Rigid geometric alignment methods (see Simeon et al.)

2. **ANOVA Methods**: Test for group differences while respecting SPD geometry
  - Fréchet ANOVA using Riemannian distances (see Muller et al.)
  - Riemannian ANOVA (closer to classical MANOVA, see Escobar, Harezlak.)
    - Wilks' lambda and Pillai's trace statistics on the tangent space  
    - Bootstrap inference for robust significance testing

By integrating harmonization and ANOVA within a Riemannian framework, `riemstats` enables rigorous statistical analysis of multi-site connectome data while preserving the geometric properties that make these matrices meaningful representations of brain connectivity.

# Why Standard Methods Fail for Connectome Data

Brain connectivity matrices (connectomes) are symmetric positive definite (SPD) matrices - they have special mathematical properties that make standard statistical methods inappropriate. This section explains why ordinary ComBat and ANOVA fail for SPD data and why specialized Riemannian methods are necessary.

## The Problem with Treating SPD Matrices as Vectors

Standard methods like ordinary ComBat and ANOVA treat matrices as if they were simple vectors in Euclidean space. This approach has critical flaws:

### 1. The Averaging Problem

```{r averaging-problem, eval=FALSE}
# Consider two correlation matrices from different subjects
# Standard averaging (element-wise mean) can produce:
# - Matrices that are no longer positive definite
# - Invalid correlation values (outside [-1, 1])
# - Loss of the geometric relationships between brain regions
```

When you average SPD matrices element-wise (as ordinary methods do), the result may not even be a valid SPD matrix. This is because SPD matrices don't form a flat vector space - they live on a curved manifold.

### 2. Distance Distortions

Standard Euclidean distance between matrices ignores their geometric structure:
- Two very similar connectivity patterns might appear far apart
- Two different patterns might appear close
- Statistical tests based on these distances give misleading results

### 3. Site Effect Removal Violations

Ordinary ComBat removes site effects by:
1. Treating each matrix element independently
2. Applying location/scale adjustments element-wise
3. Ignoring the constraint that results must be SPD

This can produce "harmonized" data that:
- No longer represents valid connectivity matrices
- Has lost important geometric relationships
- Contains negative eigenvalues (mathematically invalid for correlations)

## Why Ordinary ANOVA Fails

Standard ANOVA assumes data lies in Euclidean space where:
- The mean is computed by simple averaging
- Distances are measured with straight lines
- Variance has the same meaning everywhere

For SPD matrices, these assumptions break down:

```{r anova-failures, eval=FALSE}
# Problems with ordinary ANOVA on SPD matrices:
# 1. Group means may not be valid SPD matrices
# 2. Between-group distances are incorrectly measured
# 3. Test statistics don't account for manifold curvature
# 4. Type I error rates can be inflated or deflated
```

## The Riemannian Solution

Riemannian methods in `riemstats` solve these problems by:

1. **Proper Averaging**: Using the Fréchet mean, which always produces valid SPD matrices
2. **Correct Distances**: Measuring distances along the manifold (geodesics) rather than through Euclidean space
3. **Geometry-Preserving Harmonization**: Removing site effects while maintaining SPD structure
4. **Valid Statistical Tests**: ANOVA methods that account for manifold curvature

The key insight is that SPD matrices form a Riemannian manifold - a curved geometric space. By respecting this geometry, `riemstats` ensures:
- All operations produce valid SPD matrices
- Statistical tests have correct Type I error rates
- Biological signals are preserved during harmonization
- Results are interpretable as actual brain connectivity patterns

